---
title: "GWAS training session"
output: html_notebook
authors:
  - name: "Dr. Manuel Gentiluomo"
    affiliation: "University of Pisa"
    url: "https://github.com/mgentiluomo"
    email: "manuel.gentiluomo@unipi.it"
  - name: "Riccardo Farinella"
    affiliation: "University of Pisa"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---


🧬 Array Data Preparation: From Raw Genotypes to PLINK FormatBefore commencing any GWAS analysis, it's crucial to understand the origin and structure of the genetic data. This data is typically generated by Genotyping Array technologies (such as Illumina or Affymetrix) which detect genetic variants (the genotype) at thousands or millions of known genomic locations (SNPs).

1. 🔠 How Array Data (Raw Data) is Generated and Composed
A Genotyping Array is a chip that allows us to simultaneously test many Single Nucleotide Polymorphisms (SNPs).
Generation:
DNA from each individual is hybridised to the chip. The array software (e.g., Illumina's GenomeStudio) reads the signal intensity for the two possible alleles (genetic variants, e.g., A vs. G) at each SNP position.
Composition (Raw Genotype Data):
The result isn't immediately in a PLINK-readable format, but is usually a text file containing: 
- Sample Identifier (Sample ID): The individual being analysed.
- SNP Identifier (SNP ID): The tested genetic location.
- Raw Allelic Genotype: The raw genotype reading for that sample and SNP (e.g., AA, AG, GG, or sometimes in A/T, C/G format, etc.).

https://unipiit-my.sharepoint.com/:f:/g/personal/a035503_unipi_it/EvcItWiDOsRAn38jZlsxlZQBfta4Ifl0Y-W88ngGI7CZZQ?e=73j4pV

2. ➡️ Converting to PLINK FormatPLINK software is the de facto standard tool for GWAS analysis, but it requires data to be in a specific format to operate efficiently. This format is known as the PLINK Binary format (.bed, .bim, .fam). 
Extension | File | NameContent
.fam | Pedigree/Phenotype | Information about individuals (Family ID, Individual ID, Sex, Phenotype Status).
.bim | Genetic Map | Information about the SNPs (Chromosome, Physical Position, Alleles).
.bed | Genotype Data| The actual binary genotypes (AA, AG, GG, etc.) compressed into an efficient format for PLINK reading.

The Step from Array Matrix to PLINK:
1) Cleaning and Re-formatting: The raw output file must first be cleaned and converted into an intermediate text format (sometimes called PLINK Long Format or TPED/TFAM format).
2) PLINK Conversion: The command plink --make-bed is then used on the text format to generate the three essential binary files (.bed, .bim, and .fam).

Our entire training session will commence using files already in the PLINK Binary format.


############################################################ #############################
🛠️ Environment Setup and Prerequisites
Before starting the analysis pipeline, please follow these steps to organize your files and ensure you have all the necessary software installed. This organization is critical for the reproducibility of the entire course.

1. Folder Setup
Create the Main Project Folder: On your Desktop, create a new folder named Class_folder.

Save the Notebook: Place this R Markdown Notebook (.Rmd file) directly inside the Class_folder/.

Data Placement: All subsequent data files (PLINK data, phenotype files, etc.) that you download will be placed inside the Class_folder/ as well, usually in the 01_Database/ directory, which will be created automatically.

Your final folder structure should look like this initially:

Desktop/
└── Class_folder/
    ├── Your_GWAS_Notebook.Rmd (This file)
    └── (Other files will be created or placed here)

2. Required Software
You must have the following core software installed and accessible on your system:

R and RStudio: You should have the latest versions of R and the RStudio IDE installed. RStudio is required to run the R Markdown Notebook effectively.

PLINK: The entire Quality Control (QC) and Association Analysis pipeline relies heavily on the command-line genetic analysis tool, PLINK.

Action: You need to download the appropriate executable file for your operating system (Windows, macOS, or Linux) from the official PLINK website.

Accessibility: Ensure that the PLINK executable is either saved directly inside your Class_folder/ or, ideally, added to your system's PATH so that the system("plink ...") commands in this notebook can execute it from anywhere.

(PLINK installation and usage details will be covered in the next sections or in your accompanying slides.)

############################################################ #############################


💻 R Code Explanation: Package Installation and Setup
This specific code chunk is crucial for ensuring a smooth start to your GWAS training session. Its primary goal is to check, install, and load all the necessary R packages required for the downstream data analysis, manipulation, and visualisation.

```{r #1.a - Installation of needed packages}
# Installation of packages
packages <- c("base", "readxl", "readr", "dplyr", "ggplot2", "car", "utils", "qqman", "GGally", "ggfortify")
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
rm(packages,pkg)
```

A fundamental step in reproducible analysis is ensuring your script knows where to find the input data (data/ folder) and where to save the results (results/ folder).

When working with R Markdown, the working directory often defaults to the location of the .Rmd file itself. The following code automatically detects the path of the currently running notebook, regardless of whether you are executing it interactively in RStudio or rendering it to HTML/PDF.


Move this notebook inside the working folder
in our case is Class_folder/

```{r #1.b1 - Set of working directories}
#This chunk uses specialized functions to determine the exact folder where this notebook file resides.

# Check if the script is running interactively or being rendered
if (interactive()) {
  # Get the directory of the current R Markdown file in RStudio
  library(rstudioapi)
  notebook_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
} else {
  # Get the directory of the file during rendering
  notebook_dir <- dirname(normalizePath(knitr::current_input()))
}



# Set the working directory to the notebook's directory
setwd(notebook_dir)

# Define the main directory as the current working directory
main <- paste0(getwd(), "/")

# Print the directory for verification
print(notebook_dir)

```

This code chunk sets the local working directory and then defines a structured directory hierarchy to organise all intermediate files, data, and results generated throughout the multi-stage GWAS pipeline. This organisation is crucial for reproducibility and easy navigation.

Code Explanation: Defining Paths and Creating Folders
I. Establishing the Main Root
II. Defining the Pipeline Subdirectories
III. Creating the Folders


```{r #2 Plink}
# PLINK command setup (PLINK 2 and PLINK 1.9)

# 1. Retrieve system information
os_name <- Sys.info()[["sysname"]]

# 2. Define PLINK2 and PLINK 1.9 commands based on the operating system
if (os_name == "Windows") {
  # On Windows, executables normally end with .exe
  plink_command     <- "plink2.exe"  # PLINK 2 (already downloaded in the QC lesson)
  plink_command_old <- "plink.exe"   # PLINK 1.9 (will be added by a later chunk)
  
} else if (os_name %in% c("Darwin", "Linux")) {
  # 'Darwin' is the macOS kernel
  # On macOS/Linux we usually run executables from the current folder with "./"
  plink_command     <- "./plink2"    # PLINK 2 (already present from the QC lesson)
  plink_command_old <- "./plink"     # PLINK 1.9 (will be added by a later chunk)
  
} else {
  stop("Unsupported or unidentified operating system.")
}

# 3. Print the result for verification
cat("Detected operating system:        ", os_name, "\n")
cat("PLINK 2 command to be used:       ", plink_command, "\n")
cat("PLINK 1.9 (old) command to be used:", plink_command_old, "\n")

# --- Example usage (optional checks) ---

cat("\n--- Testing PLINK 2 (if available) ---\n")
system(paste(plink_command, "--version"))

cat("\n--- Testing PLINK 1.9 (if available) ---\n")
system(paste(plink_command_old, "--version"))

print("End of chunk 2")
```

```{r download_and_prepare_plink1.9}
# Download and prepare PLINK 1.9 (old PLINK)
# This chunk does NOT modify or overwrite plink2.
# It simply adds the plink 1.9 executable alongside it.

# Use the existing notebook directory if defined; otherwise fall back to the current working directory
if (!exists("notebook_dir")) {
  notebook_dir <- getwd()
}

# Detect the operating system (reuse os_name if it already exists)
if (!exists("os_name")) {
  os_name <- Sys.info()[["sysname"]]
}

# Define the PLINK 1.9 download URLs (update these if a newer build is released)
plink_url_mac     <- "https://s3.amazonaws.com/plink1-assets/plink_mac_20250819.zip"
plink_url_windows <- "https://s3.amazonaws.com/plink1-assets/plink_win64_20250819.zip"

# Choose the correct URL based on the operating system
if (os_name == "Windows") {
  plink_url <- plink_url_windows
} else if (os_name == "Darwin") {
  plink_url <- plink_url_mac
} else {
  stop("This PLINK 1.9 download script currently supports only Windows and macOS.")
}

# Derive the ZIP file name from the URL
file_name <- basename(plink_url)

# Define the destination path for the ZIP file
dest_file <- file.path(notebook_dir, file_name)

# Download the ZIP file
message("⬇️ Downloading PLINK 1.9 from: ", plink_url)
download.file(url = plink_url, destfile = dest_file, mode = "wb")

# Confirm that the ZIP file exists
if (!file.exists(dest_file)) {
  stop("Download failed: ZIP file not found at ", dest_file)
}

# Create a folder to extract the ZIP contents (named after the ZIP file, without .zip)
extract_dir <- file.path(
  notebook_dir,
  tools::file_path_sans_ext(basename(dest_file))
)

if (!dir.exists(extract_dir)) dir.create(extract_dir)

# Unzip the contents into extract_dir
unzip(zipfile = dest_file, exdir = extract_dir)
message("Extracted: ", basename(dest_file), " → ", extract_dir)

# Find any file inside the extracted folder that contains "plink" in its name
# (this will typically include "plink" / "plink.exe", but NOT your existing plink2)
plink_files <- list.files(
  path = extract_dir,
  pattern = "plink",
  full.names = TRUE,
  recursive = TRUE
)

# Copy each found PLINK 1.9 file back into the main notebook directory
for (pf in plink_files) {
  file.copy(from = pf, to = notebook_dir, overwrite = TRUE)
  message("Copied: ", basename(pf), " → ", notebook_dir)
}

# On macOS/Linux, ensure the plink 1.9 binary is executable
if (os_name != "Windows") {
  plink_old_path <- file.path(notebook_dir, "plink")
  if (file.exists(plink_old_path)) {
    Sys.chmod(plink_old_path, mode = "0755")
    message("Set executable permissions on: ", plink_old_path)
  }
}

# List all PLINK-related files now present in the main notebook directory
message("PLINK-related files in the notebook directory:")
print(list.files(notebook_dir, pattern = "plink", full.names = TRUE))

rm(plink_url,plink_url_mac,plink_url_windows)
```

Statistical analysis
Association analysis 
In GWAS for diseases, the goal is to model the association of SNPs with diseases. Since the outcome in this case takes two values (“disease” or “no disease”), statistical modelling methods for binary outcomes are suited for this task.
The standard approach in this context is to use logistic regression. The logistic regression model assumes that the probability of the binary outcome is related to a linear combination of genetic predictors, adjusted for potential confounding factors.

Note: Since the post-imputation files are too large and will require too long to be processed we will use the pre-imputed files. 

Note2: When using a mac add  ./ before plink
 
PLINK code for logistic regression adjusted only for sex:

plink --bfile DB --logistic --ci 0.95 --sex --hide-covar --out Logistic_unadjusted

```{r}
system(paste(
  plink_command_old,
  " --bfile 03_PCA/DB_EUR",
  " --logistic",
  " --ci 0.95",
  " --sex",
  " --hide-covar",
  " --out 06_Statistical_analysis/Logistic_unadjusted"
))
```

 
You will obtained a file as output named : namefile.assoc.logistic
 
plink.assoc.logistic
depending on the phenotype/command used. The basic format of the output file is:
     CHR       Chromosome
     SNP       SNP identifier
     BP        Physical position (base-pair)
     A1        Tested allele (minor allele by default) 
     TEST      Code for the test (see below)
     NMISS     Number of non-missing individuals included in the analysis
     BETA/OR   Regression coefficient
     STAT      Coefficient t-statistic 
     P         Asymptotic p-value for t-statistic
For the additive effects of SNPs, the direction of the regression coefficient represents the effect of each extra minor allele (i.e. a positive regression coefficient means that the minor allele increases risk/phenotype mean). If the --beta command is added along with --logistic, then the regression coefficients rather than the odds ratios will be returned.

```{r}
Logistic_unadjusted <- read.table("06_Statistical_analysis/Logistic_unadjusted.assoc.logistic", header=TRUE, quote="", comment.char="")
```


PLINK code for logistic regression adjusted only for sex and other covariates:
```{r COVAR}
# Import study population covariates, such as sex, phenotype (likely already updated in the .fam file), age, BMI, etc.
COV <- read.delim("02_Database/general_covar.txt", quote = "")
```
 
plink --bfile DB --logistic --covar DB_covariates.txt --covar-name AGE --hide-covar --ci 0.95 --out Logistic_adjusted

```{r}
system(paste(
  plink_command_old,
  " --bfile 03_PCA/DB_EUR",
  " --logistic",
  " --covar 02_Database/general_covar.txt",
  " --covar-name Age",
  " --ci 0.95",
  " --sex",
  " --hide-covar",
  " --out 06_Statistical_analysis/Logistic_adjusted"
))
```

 
The
--covar-name or --covar-number commands can be used to select a subset of all
covariates in the file.
plink
--file mydata --covar c.txt --covar-number 2,4-6,8
or
plink
--file mydata --covar c.txt --covar-name AGE,BMI-SMOKE,ALC

```{r}
Logistic_adjusted <- read.table("06_Statistical_analysis/Logistic_adjusted.assoc.logistic", header=TRUE, quote="", comment.char="")
```

To account for population stratification, we will use the PCA files generated before and create a new covariate file. 

```{r --pca}
# SECTION 6: Cryptic relatedness check

# Step 1: Prune SNPs for PCA analysis.
system(paste(plink_command, " --bfile 02_QC/QC9_hwe --indep-pairwise 50 5 0.2 --out 03_PCA/Pruned_list_to_PCA"))


# STEP 2: Extraction of independent SNPs
system(paste(plink_command," --bfile 02_QC/QC9_hwe --extract 03_PCA/Pruned_list_to_PCA.prune.in --pca --out 06_Statistical_analysis/PCA"))

# Load PCA eigenvectors
PCA <- read.delim("06_Statistical_analysis/PCA.eigenvec", quote = "",)
num_pca_columns <- ncol(PCA) - 2
colnames(PCA) <- c("FID", "IID", paste0("PC", 1:num_pca_columns))

COVAR <- merge(COV,PCA, all = T)

write.table(COVAR, "06_Statistical_analysis/COVAR.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```


# Read the existing covariate file
covariate_file <- read.table("covariates.txt", header = TRUE, sep = "\t")
 
# Read the PC scores file
pc_scores_file <- read.table("pc_scores.txt", header = TRUE, sep = "\t")
 
# Merge the covariate file and PC scores file based on FID and IID
merged_data <- merge(covariate_file, pc_scores_file, by = c("FID", "IID"))
 
# Write the merged data to a new covariate file
write.table(merged_data, "covariates_with_pc.txt", sep = "\t", quote = FALSE, row.names = FALSE)
 
PLINK code for logistic regression adjusted also for principal components:
 
plink --bfile DB --logistic --hide-covar --ci 0.95 --sex --covar DB_covariates.txt --covar-name AGE,PC1-PC10 --out Logistic_adj_age_PC10

```{r}
system(paste(
  plink_command_old,
  " --bfile 03_PCA/DB_EUR",
  " --logistic",
  " --covar 06_Statistical_analysis/COVAR.txt",
  " --covar-name Age PC1-PC10",
  " --ci 0.95",
  " --sex",
  " --hide-covar",
  " --out 06_Statistical_analysis/Logistic_adj_age_PC10"
))
```

```{r}
Logistic_adj_age_PC10 <- read.table("06_Statistical_analysis/Logistic_adj_age_PC10.assoc.logistic", header=TRUE, quote="", comment.char="")
```
 
Where PC1, … , PC10 are the ten first principal components of the SNP matrix.


To calculate allele frequencies
plink --bfile DB --freq --out Allele_Freq

```{r}
system(paste(
  plink_command_old,
  " --bfile 03_PCA/DB_EUR",
  " --freq",
  " --out 06_Statistical_analysis/Allele_Freq"
))

Freq <- read.table("06_Statistical_analysis/Allele_Freq.frq", header=TRUE, quote="", comment.char="")
```


Covariates selection and correlation matrix 
Why should we perform covariate selection?
The choice of which covariates to include in a Genome-Wide Association Study (GWAS) is important because it affects the ability to detect true association signals, and to correct for confounders avoiding false positive results. Commonly used covariates are age, sex, as well as an arbitrary number of Principal Components (PCs) used to adjust for population structure.
Why correlation matrix?
A correlation matrix is a statistical technique used to evaluate the relationship between two or more variables in a data set. The concept is very similar to that of a covariance matrix, which has been previously used for PCA. By using a correlation matrix, we can evaluate the relationship between two variables:
If relationship is 1, then the relationship is strong
If relationship is 0, then the relationship is neutral
If relationship is -1, then the relationship is strong

```{r Covariate tertiles and dummies}
library(ggplot2)
library(GGally)
library(dplyr)

### 1. Correlation matrix (just as in your original code)
GGally::ggpairs(COVAR, columns = 5:9)

### 2. Create a filtered dataset to compute tertiles of Pack.Years among smokers
DB_covar_filtered <- COVAR %>%
  filter(Pack.Years != 0,
         !is.na(Pack.Years))

### 3. Compute tertiles of Pack.Years (only among smokers)
tertiles <- quantile(DB_covar_filtered$Pack.Years,
                     probs = seq(0, 1, 1/3),
                     na.rm = TRUE)

tertiles

### 4. Dummy variable creation based on Pack.Years
# never  = never smokers (Pack.Years == 0)
# low    = first tertile among smokers
# middle = second tertile among smokers
# high   = third tertile among smokers

COVAR$never  <- ifelse(COVAR$Pack.Years == 0 | is.na(COVAR$Pack.Years), 1, 0)

COVAR$low    <- ifelse(COVAR$Pack.Years > 0 &
                         COVAR$Pack.Years <= tertiles[2], 1, 0)

COVAR$middle <- ifelse(COVAR$Pack.Years > tertiles[2] &
                         COVAR$Pack.Years <= tertiles[3], 1, 0)

COVAR$high   <- ifelse(COVAR$Pack.Years > tertiles[3], 1, 0)

### 5. Drop original continuous variables if you do not want them in the model
COVAR <- dplyr::select(COVAR, -c(Pack.Years, BMI, pheno))

### 6. Write the covariate file to disk
write.table(
  COVAR,
  file      = "06_Statistical_analysis/DB_covar_dummy.txt",
  sep       = "\t",
  quote     = FALSE,
  col.names = TRUE,
  row.names = FALSE
)
```



New association analysis


Data visualisation
 
Once you've completed the separate logistic regression analyses for each SNP and computed their corresponding p-values, the next intuitive step would be data visualisation. One of the most common and insightful ways to display GWAS data is through Manhattan plots.
Creating a Manhattan plot involves plotting each SNP's genomic position (from all the chromosomes) on the x-axis and the corresponding -log10(p-value) on the y-axis.The "skyline" of this plot contains peaks that represent SNPs with lower p-values (higher -log10(p-value)), indicating a potential genetic association with the disease under investigation.
 
Alongside Manhattan plots, another tool in visualizing GWAS results is the Quantile-Quantile (QQ) plot. QQ plots offer a graphical means to assess if the p-values from the GWAS deviate from the expected distribution under the null hypothesis, which assumes no association between SNPs and the disease.
Creating a QQ plot involves plotting the observed p-values against their expected values under the null hypothesis. On the x-axis, we plot the theoretical or expected -log10(p-values) under the null hypothesis, while on the y-axis, we plot the observed -log10(p-values) from the GWAS.
Under the null hypothesis of no associations, the points should approximately follow the 45-degree line (also known as the line of equality), indicating that the observed distribution of p-values matches the expected distribution. However, the presence of true genetic associations can cause a deviation from this line.
Specifically, if there are SNPs associated with the disease, we would expect to see an upward departure from the line of equality at the right end of the plot, representing observed p-values that are more significant than expected by chance. This 'tail' of the QQ plot helps identify the presence of true associations.
 
 
We are going to use R/Rstudio to create both graphical representations.
 
Codes for Mannhattan plot generation:
```{r}
library(qqman)

png("06_Statistical_analysis/manhattan.png", height = 2000, width = 3000, units = "px", res = 400, )
print(manhattan(Logistic_adj_age_PC10, annotatePval =
5e-05,chr="CHR", bp="BP", snp="SNP",
p="P", ylim = c(0, 13),chrlabs = c(1:22), suggestiveline =
-log10(1e-05), genomewideline = -log10(5e-08)))
dev.off()
```
Note: To make the code compatible with a Mac, you need to adjust the file path format. On a Mac, the file path format typically starts with / and uses forward slashes (/) as directory separators.

Your png file will be ready in the specified path and will look like this:


```{r Codes for qqplot generation:}
library(qqman)

png("06_Statistical_analysis/qqplot_l.png", width = 2000, height = 2000, units="px", res=400, type="cairo")

lambda<-median(qchisq(1-na.omit(Logistic_unadjusted$P),1))/qchisq(0.5,1)

qq(Logistic_unadjusted$P)

text(1,6, paste("lambda","=",  signif(lambda, digits = 3)) )
dev.off()
```

Your png file will be ready in the specified path and will look like this:
 
```{r Code to generate a table of SNP’s association filtered for p-values (p-value<5E-5)}
#Code to generate a table of SNP’s association filtered for p-values (p-value<5E-5)
 
Results5e5 <- subset(Logistic_adj_age_PC10, (Logistic_adj_age_PC10$P < 0.00005))
 
Results5e5 <- merge(Results5e5,Freq)
 
Results5e5$NCHROBS =NULL
 
write.table(Results5e5,"06_Statistical_analysis/Results5e5.txt",
row.names=F, col.names=T, sep="\t", quote=F)
```


```{r Manhattan plot colour}
library(ggplot2)
library(dplyr)

### Save high-resolution PNG
png(
  filename = "06_Statistical_analysis/ManhattanC.png",
  width = 4500,      # wider = better SNP separation
  height = 1800,     # good vertical size for y-axis labels
  units = "px",
  res = 350          # high resolution
)

don <- Logistic_unadjusted %>% 
  
  # Compute chromosome size
  group_by(CHR) %>% 
  summarise(chr_len=max(BP)) %>% 
  
  # Calculate cumulative position of each chromosome
  mutate(tot=cumsum(chr_len)-chr_len) %>%
  select(-chr_len) %>%
  
  # Add this info to the initial dataset
  left_join(gwasResults, ., by=c("CHR"="CHR")) %>%
  
  # Add a cumulative position of each SNP
  arrange(CHR, BP) %>%
  mutate( BPcum=BP+tot)

axisdf = don %>%
  group_by(CHR) %>%
  summarize(center=( max(BPcum) + min(BPcum) ) / 2 )
#Ready to make the plot using ggplot2:

ggplot(don, aes(x=BPcum, y=-log10(P))) +
    
    # Show all points
    geom_point( aes(color=as.factor(CHR)), alpha=0.8, size=1.3) +
    scale_color_manual(values = rep(c("grey", "skyblue"), 22 )) +
    
    # custom X axis:
    scale_x_continuous( label = axisdf$CHR, breaks= axisdf$center ) +
    scale_y_continuous(expand = c(0, 0) ) +     # remove space between plot area and x axis
  
    # Custom the theme:
    theme_bw() +
    theme( 
      legend.position="none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )

dev.off()
```


